---
title: "Workshop 8 TEMPLATE"
subtitle: "12 PM"
author: "Megan Wien"
date: "04/22/26"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(here("data", "Valliere_etal_EcoApps_Data.xlsx"),sheet = "First Harvest")


```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating new object
drought_exp_clean <- drought_exp |> 
  # cleaning column names - lowercase with no spaces (_)
  clean_names() |> 
  # renaming species names
  mutate(species_name = recode_values(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # renaming water treatments
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # assigning order of variables
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting specific columns of interest
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  
```{r}
#| label: checking-structure
str(drought_exp_clean)
```




Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: ten-rows
# clean data frame displaying 10 rows
slice_sample(drought_exp_clean, n = 10)

```




# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 
```{r}
#| label: scatterplot

ggplot(data = drought_exp_clean, mapping = aes(x = sla, y = total_g)) +
  geom_point() +
  labs(x = "SLA (cm"^2~"/g)",
       y = "Total Mass (g)",
       title = "As SLA inreases so does the total mass")

```

## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  
```{r}
#| label: jitter
ggplot(data = drought_exp_clean, mapping = aes(x = water_treatment,
                                               y = total_g,
                                               color = water_treatment)) + 
  geom_jitter(height = 0) +
  
  labs( x = "Water Treatment", y = "Total Mass (g)", title = "Well watered plants have a greater mean total mass than drought treated plants")
  
```

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

```{r}
#| label: jitter-by-species

ggplot(data = drought_exp_clean, mapping = aes(x = species_name,
                                               y = total_g,
                                               color = species_name)) + 
  geom_jitter(height = 0, width = 0.2) + 
  labs( x = "Species", y = "Total Mass (g)", title = "Each species has a different mean total mass with Grindelia camporum as the highest and Lotus scoparius
 as the lowest") +
theme(legend.position = "none") + 
  scale_color_manual(values = c("Lotus scoparius" = lotsco_col,
                                "Penstemon centranthifolius" = pencen_col, "Salvia leucophylla" = salleu_col, "Encelia californica" = enccal_col, "Stipa pulchra" = stipul_col, "Eschscholzia californica" = esccal_col, "Grindelia camporum" = gricam_col))

```



Bonus: order the x-axis by decreasing mean plant mass.

# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(total_g ~ 1, data = drought_exp_clean)

# model 1: all predictors
model1 <- lm(total_g ~ water_treatment + species_name + sla, data = drought_exp_clean)

# model 2: SLA and water treatment
model2 <- lm(total_g ~ sla + water_treatment, data = drought_exp_clean)

# model 3: SLA and species
model3 <- lm(total_g ~ sla + species_name, data = drought_exp_clean)

# model 4: water treatment and species
model4 <- lm(total_g ~ water_treatment + species_name, data = drought_exp_clean)

# model 5: SLA
model5 <- lm(total_g ~ sla, data = drought_exp_clean)

# model 6: water treatment
model6 <- lm(total_g ~ water_treatment, data = drought_exp_clean)

# model 7: species
model7 <- lm(total_g ~ species_name, data = drought_exp_clean)
```

## Model diagnostics
```{r}
#| label: checking-diagnostic
par(mfrow = c(2,2))
plot(model4)
```
Residuals are normally distributed from the qqplot
residuals are homoscedastic - residuals vs fitted/scale location
no outliers that could be influencing line of best fit

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: model-selection
AICc(model0, model1, model2, model3, model4, model5, model6, model7) |> 
  arrange(AICc)
```



Insert code to display the summary for the best model. 
```{r}
#| label: model-summary

summary(model4)

```





## Stop and think: what does this model mean?  

**What is the best model?**

The best model is model 4 - which is water treatment and species as predictors.

**How much variation in the response (total mass, in grams) does this model explain?**

72.6% of the variation in plant mass is explained by this model.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

Species influences plant mass. 
Treatment influences plant mass.
 -  The `(intercept)` is the model prediction for plant mass for Lotus scoparious in the drought stressed treatment. 
 - The `water_treatmentWell watered`: model prediction for the difference in Lotus scoparius mass between well watered and drought stressed treatments.
 - `species_namePenstemon centranthifolius`: difference between Penstemon mass in well watered treatments and the Lotus scoparius mass in the well watered treatment. 
 - `species_nameEschscholzia californica`: difference between the Eschscholzia californica mass in well watered treatment and the Lotus scoparius mass in the well watered treatment. 

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.
```{r}
#| label: model-predictions

preds <- ggpredict(model4, terms = c("water_treatment", "species_name")) |> # predictors and model object
rename(water_treatment = x, species_name = group
      )


```



# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.
```{r}
#| label: final figure

ggplot(data = drought_exp_clean, mapping = aes(x = water_treatment, y = total_g, color = water_treatment)) + 
  geom_jitter(height = 0, width = 0.1, alpha = 0.4, shape = 21) +
  geom_pointrange(data = preds, mapping = aes(x = water_treatment, y = predicted, ymin = conf.low, ymax = conf.high)) +
  facet_wrap(~ species_name) + 
  scale_color_manual(values = c("Drought stressed" = ds_col, "Well watered" = ww_col)) + 
  theme(legend.position = "none") + labs(x = "Treatment", y = "Total Mass(g)")
```

# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







